Spin-orbit-coupling-induced squeezing in the optical lattice clock

### 2018-07-19 soc_squeezing/soc_squeezing_notes.tex
moved derivation of OAT Hamiltonian to an appendix

\usepackage{dsfont} % for identity operator
\newcommand{\1}{\hat{\mathds{1}}}

@@

These are some notes about trying to induce squeezing in an optical
lattice clock by combining spin-orbit coupling and interactions.  We
assume, for now, that
\begin{enumerate*}[label=(\roman*)]
\item atoms are fermionic in nature with an alkali-earth(-like)
  electronic structure,
\item atoms are nuclear-spin-polarized, and
\item the optical lattice is quasi-one-dimensional, with tight
  (i.e.~ground-state) confinement in transverse directions.
\end{enumerate*}

@@

If the single-particle energy spacings are much greater than the
strength of inter-particle interactions, or equivalently if
$J_0\gg U$, then by the secular approximation we can neglect terms in
$H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}\ne\set{r,s}$; that
is, atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_full} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}.
\end{align}
Up to a global shift in energy, this result is equivalently
\begin{align}
  H_{\t{int}}
  = - \f14 \f{U}{L} \sum_{p,q} \v{\sigma}_q\c\v{\sigma}_p
  = - \f{U}{L} \sum_{p,q} \v s_q \c\v s_p
  = - \f{U}{L} \v S \c \v S,
\end{align}
where
\begin{align}
  \v{\sigma}_p
  \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z},
  &&
  \v s_p \equiv \p{s_p^x,s_P^y,s_p^z},
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
  \label{eq:H_int}
\end{align}
For a fixed particle number $N$, the collective spin Hamiltonian
$H_{\t{int}}$ has electronic eigenstates $\set{\ket{Sm}}$ with total
(pseudo-)spin $S$ and projection $m$ onto a quantization axis; the
corresponding energies are $\bk{Sm|H_{\t{int}}|S,m}=-\p{U/L}S\p{S+1}$.
The ground-state $S=N/2$ manifold is spanned by the Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m}
  = {N \choose N/2+m}^{-1/2} S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
\end{align}
for which we note that $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$.  Above this
ground-state manifold with an energy gap of $NU/L=\eta U$ lies the
manifold of $S=N/2-1$ states, which is spanned by the orthonormal
spin-wave states
\begin{align}
  \ket{mk}
  \equiv \ket{N/2-1,m,k}
  \equiv {N-1 \choose \p{N/2-m}\p{N/2-m+1}}^{-1/2}
  \sum_{n=1}^N e^{2\pi ikn/N} s_n^+ \ket{N/2,m-1},
\end{align}
where $k=1,2,\cdots,N-1$; $n$ indices an individual atom; and
$s_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$
\cite{swallows2011suppression}.  \note{[We should comment about or
  otherwise modify the text to reflect the fact that do not actually
  achieve $J_0\gg U$ in practice.]}

@@

In order to demonstrate spin squeezing behavior induced by spin-orbit
coupling, we fix the total particle number $N$, consider initial
states with total spin $S=N/2$, and write the total system Hamiltonian
as

@@

  H_{\t{free}} = H_{\t{int}} + H_{\t{SOC}}
  = -\f{U}{L} \v S\c\v S - \sum_n h_n s_n^z
  = -\f{U}{L} \v S\c\v S - \bar h S_z - \sum_n b_n s_n^z,

@@

where we have separated out the mean and residual effective fields

@@

  b_n \equiv h_n - \bar h.

@@

The subscript on $H_{\t{free}}$ refers to the fact that this
Hamiltonian corresponds to free, undriven evolution of (interacting)
atoms on the lattice.  The collective spin operators $\v S\c\v S$ and
$S_z$ both preserve total spin $S$ and its projection $m$ onto a
quantization axis, which makes the Dicke states a good basis for
treating $H_{\t{free}}$ (i.e.~within the $S=N/2$ manifold).  While the
residual effective field term $\sum_n b_n s_n^z$ generally takes
states outside the $S=N/2$ manifold, we can treat this spin-changing
process perturbatively when the residual effective fields $b_n$ are
small compared to the collective spin gap, i.e.~when
$\max_n\abs{b_n}\ll \eta\abs{U}$.  Such a perturbative treatment
yields an effective Hamiltonian of the form
$H_{\t{free,eff}} = H_{\t{free,eff}}^{(1)} + H_{\t{free,eff}}^{(2)}$
with

@@

  H_{\t{free,eff}}^{(1)}
  &= \sum_m \op{m} H \op{m}
  = -\f{U}{L} \v S \c \v S - \bar h S_z, \\
  &= -\sum_{\substack{m\\S<N/2}}
  \frac{\abs{\bk{Sm|H|m}}^2}{E_{Sm}^{\t{free}}-E_m^{\t{free}}} \op{m}
  = -\sum_{\substack{m\\S<N/2}}
  \frac{\abs{\sum_n b_n \bk{Sm|s_n^z|m}}^2}
  {E_{Sm}^{\t{free}}-E_m^{\t{free}}} \op{m},

@@

where $E_\psi^{\t{free}}\equiv\bk{\psi|H_{\t{free}}|\psi}$ and we have
used the fact the spin projection $m$ is conserved by $H$.  As the
individual spin operators $s_n^z$ cannot change the total spin $S$ by
more than 1, we need only consider $S=N/2-1$ spin-wave states when
evaluating the sum over $S<N/2$ in $H_{\t{free,eff}}^{(2)}$ above.
These spin-wave states couple to the Dicke states via the matrix
elements of the form (see section 2 in the supporting materials of
\cite{swallows2011suppression})

@@

  \bk{m|s_n^z|mk} = e^{2\pi i k n/N} \sqrt{\f{(N/2)^2-m^2}{N^2 (N-1)}},

@@

  = -\sum_{m,k}
  \frac{\abs{\sum_n b_n \bk{mk|s_n^z|m}}^2}{\eta U} \op{m}
  = -\sum_m \f{(N/2)^2-m^2}{N^2\p{N-1}\eta U}
  \sum_k \abs{\sum_n b_n e^{2\pi ikn/N}}^2 \op{m}.

@@

  \sum_k \abs{\sum_n b_n e^{2\pi ikn/N}}^2
  = \sum_{k=1}^{N-1} \sum_{n,n'} b_n b_{n'} e^{2\pi ik\p{n-n'}/N}

@@

  \sum_k \abs{\sum_n b_n e^{2\pi ikn/N}}^2

@@

  H_{\t{free,eff}}
  = -\f{U}{L} \v S \c \v S - \bar h S_z + \chi S_z^2,

@@

This effective Hamiltonian preserves total spin $S$, which means that
the $\v S\c\v S$ term contributes only a constant energy shift which
we can neglect.  Furthermore, at temperatures $T\gg J_0$ the mean
effective field $\bar h=0$, while for $T\sim J_0$ the collective
rotation due to the $S_z$ term can be neglected by use of a rotating
frame.  The effective Hamiltonian therefore simplifies to the one-axis
twisting Hamiltonian
  H_{\t{OAT}}^z = \chi S_z^2.
  \label{eq:H_OAT}
An example set of parameters and values for one-axis twisting via
$H_{\t{OAT}}^z$ with strontium-87 is provided in Table
\ref{tab:parameters}, along with the optimal one-axis twisting
squeezing time
$t_{\t{opt}}^{\t{OAT}}\sim N^{-2/3}\chi^{-1}\sim\eta
N^{1/3}=N^{4/3}/L$.  Lattice depths are provided in units of the
lattice recoil energy $E_R$.

While one-axis twisting provides a spin fluctuation noise floor which
scales with atom number as $\sim1/N^{2/3}$, it is possible to beat
this limit and achieve a noise floor scaling as $\sim1/N$ via
so-called two-axis twisting.  We can transform the one-axis twisting
Hamitlonian in \eqref{eq:H_OAT} into an effective two-axis twisting
Hamiltonian by using a pulsed drive sequence developed in
ref.~\cite{liu2011spin}.  The necessary drive sequence requires the
pulsed application of a clock laser, which turns on a collective spin
Hamiltonian
\begin{align}
  H_{\t{clock}} = -\Omega \sum_n s_n^x = -\Omega S_x.
\end{align}
If $\abs{\Omega}\gg N\chi$, then we can apply $H_{\t{clock}}$ for
short times $\theta/\abs{\Omega}\ll\p{N\chi}^{-1}$ (during which we
can neglect $H_{\t{eff}}$) to realize collective rotations of the form
$\exp\p{\pm i\theta S_x}$.  Thus
\begin{enumerate*}[label=(\roman*)]
\item acting with $\exp\sp{i\p{\pi/2}S_x}$,
\item waiting for a time $2\tau$,
\item acting with $\exp\sp{-i\p{\pi/2}S_x}$, and
\item waiting for a time $\tau$
\end{enumerate*}
realizes the unitary
\begin{align}
  U_1 = \exp\p{-i\tau\chi S_z^2} \exp\p{-i\f{\pi}{2}S_x}
  \exp\p{-i2\tau\chi S_z^2} \exp\p{i\f{\pi}{2}S_x}
  = \exp\p{-i\tau\chi S_z^2} \exp\p{-i2\tau\chi S_y^2}.
\end{align}
If $\tau\chi\ll1$, then
\begin{align}
  U_1 \approx \exp\sp{-i\tau\chi\p{S_z^2 + 2S_y^2}}
  = \exp\sp{-i\tau\chi\p{S^2 + S_y^2 - S_x^2}}
  \simeq \exp\sp{-i\tau\chi\p{S_y^2 - S_x^2}},
\end{align}
where $\simeq$ denotes equality up to an overall phase, which holds
above because $S^2$ is conserved by all dynamics.  Repeating the above
sequence $n$ times for a time $t=3n\tau$ thus results in the unitary
\begin{align}
  U = \exp\sp{-in\tau\chi\p{S_y^2 - S_x^2}}
  = \exp\sp{-i\p{t/3}\chi\p{S_y^2 - S_x^2}},
\end{align}
which is equivalently generated by evolution for a time $t$ under the
two-axis twisting Hamiltonian
\begin{align}
  H_{\t{TAT}}^{y,x} = \f{\chi}{3}\p{S_y^2 - S_x^2}.
  \label{eq:H_TAT_pulse}
\end{align}
Despite the factor of 1/3 in the effective two-axis twisting strength,
the optimal squeezing time for this two-axis twisting Hamiltonian is
generally close to that for one-axis twisting with $H_{\t{OAT}}^z$ in
\eqref{eq:H_OAT} \cite{liu2011spin}.

\begin{figure}
  \centering
  \includegraphics{squeezing_comparison.pdf}
  \caption{Comparison of squeezing through $H_{\t{OAT}}^z$ and
    $H_{\t{TAT}}^{z,x}$ via the parameters in Table
    \ref{tab:parameters} and the continuous two-axis twisting protocol
    with $\omega=27\times2\pi$ Hz and $\beta=0.906$.}
  \label{fig:squeezing_comparison}
\end{figure}

\section{Effective spin squeezing in the presence of a drive}
\label{sec:effective_squeezing_drive}

Suppose we have a Hamiltonian of the form
\begin{align}
  H = H_0 + V,
  &&
  H_0 = - \xi\v S\c\v S,
  &&
  V = - \bar h S_z - \Omega S_x - \sum_n b_n s_n^z,
\end{align}
where the effective fields $b_n$ are mean-zero, and we consider
initial states in the ground-state manifold $\G_0$ of $H_0$ (i.e.~with
total spin $S=N/2$).  If the largest eigenvalue of $V$ is smaller in
magnitude than half of the collective spin gap $N\xi$, or the energy
gap under $H_0$ between $\G_0$ and its orthogonal complement $\E_0$,
then we can formally develop a perturbative treatment of the action of
$V$ on $\G_0$.  Such a treatment yields an effective Hamiltonian
$H_{\t{eff}}=\sum_pH_{\t{eff}}^{(p)}$ on $\G_0$, where
$H_{\t{eff}}^{(p)}$ is order $p$ in $V$.  Letting $\P_0$ ($\Q_0$) be a
projector onto $\G_0$ ($\E_0$) and $X$ denote any operator on
$\G_0\cup\E_0$ (i.e.~the entire Hilbert space), we define the
superoperators
\begin{align}
  \D X \equiv \P_0 X \P_0 + \Q_0 X \Q_0,
  &&
  \O X \equiv \P_0 X \Q_0 + \Q_0 X \P_0,
\end{align}
which select the diagonal ($\D$) and off-diagonal ($\O$) parts of $X$
with respect to $\G_0$ and $\E_0$, and
\begin{align}
  \L X \equiv \sum_{\alpha,\beta}
  \f{\op{\alpha}\O X\op{\beta}}{E_\alpha-E_\beta},
  &&
  \t{where}
  &&
  H_0 = \sum_\alpha E_\alpha \op\alpha.
\end{align}
The first few terms in the expansion of the effective Hamiltonian are
then, as derived in ref.~\cite{bravyi2011schrieffer},
\begin{align}
  H_{\t{eff}}^{(0)} = \P_0 H_0 \P_0,
  &&
  H_{\t{eff}}^{(1)} = \P_0 V \P_0,
  &&
  H_{\t{eff}}^{(2)} = -\f12 \P_0 \sp{\O V,\L V} \P_0.
\end{align}
Letting $\B_0\p{\E_0}$ denote an eigenbasis of $H_0$ for $\E_0$ and
adding a constant shift to $H_0$ such that $E_\psi=0$ for
$\ket\psi\in\G_0$, we additionally define
\begin{align}
  \I \equiv \sum_{\ket\alpha\in\B_0\p{\E_0}} \f{\op\alpha}{E_\alpha},
\end{align}
which sums over projections onto excited states with corresponding
energetic suppression factors, and expand
\begin{align}
  \L X = \O\p{\L X}
  = \Q_0 \L X \P_0 + \P_0 \L X \Q_0
  = \I X \P_0 - \P_0 X \I.
\end{align}
The expression for $H_{\t{eff}}^{(2)}$ then simplifies to
\begin{align}
  H_{\t{eff}}
  = -\f12 \P_0 \p{\sp{\O V, \I V \P_0} - \sp{\O V, \P_0 V \I}} \P_0
  = -\P_0 V \I V \P_0.
\end{align}
\note{[MP: much of the text above is essentially copied from an
  appendix of the theory paper for multi-body interactions.  As soon
  as that paper gets posted on arXiv, we can cite it instead.]}  It
follows that
\begin{align}
  H_{\t{eff}}^{(0)} = 0,
  &&
  H_{\t{eff}}^{(1)} = -\bar h S_z - \Omega S_x,
\end{align}
and, in terms of the $S=N/2$ Dicke states $\ket{m}\equiv\ket{N/2,m}$
and the $S=N/2-1$ spin-wave states $\ket{mk}\equiv\ket{N/2-1,mk}$,
\begin{align}
  H_{\t{eff}}^{(2)}
  = - \f1{N\xi} \sum_{m,k} \abs{\sum_nb_n\bk{mk|s_n^z|m}}^2 \op{m},
  \label{eq:general_H_eff_2}
\end{align}
where we used the fact that the collective spin operators $S_j$
conserve total spin $S$, while the individual spin operators $s_n^z$
conserve the projection $m$ of collective spin onto a quantization
axis.  The second-order effective Hamiltonian $H_{\t{eff}}^{(2)}$ in
\eqref{eq:general_H_eff_2} is identical to that derived in Section
\ref{sec:OAT}, which implies
\begin{align}
  H_{\t{eff}}^{(2)} = \chi S_z^2,
  &&
  \chi \equiv \f{\tilde h^2}{N\p{N-1}\xi},
  &&
  \tilde h^2 \equiv \f1N \sum_n b_n^2,
\end{align}
and in turn
\begin{align}
  H_{\t{eff}} = -\bar h S_z - \Omega S_x + \chi S_z^2.
\end{align}

>>>>

\usepackage{dsfont} % for identity operator
\newcommand{\1}{\mathds{1}}

@@

These are some notes about trying to induce spin squeezing in an
optical lattice clock by combining spin-orbit coupling, interactions,
and external drives.  We assume, for now, that
\begin{enumerate*}[label=(\roman*)]
\item atoms are fermionic in nature with an alkali-earth(-like)
  electronic structure,
\item atoms are nuclear-spin-polarized, and
\item the optical lattice is quasi-one-dimensional, with tight
  (i.e.~ground-state) confinement in transverse directions.
\end{enumerate*}

@@

If the single-particle energy spacings are large compared to the
strength of inter-particle interactions, or equivalently if
$J_0\gtrsim U$, then by the secular approximation we can neglect terms
in $H_{\t{int}}$ which do not conserve the sum of single-particle
energies.  By solving this single-particle energy conservation
condition, one finds that we must have $\set{p,q}\ne\set{r,s}$; that
is, atoms in different clock states can only interact via
\begin{enumerate*}[label=(\roman*)]
\item direct density-density terms, and
\item terms which exchange their momenta.
\end{enumerate*}
The surviving terms in \eqref{eq:H_int_full} are thus
\begin{align}
  H_{\t{int}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{p,\e}^\dag \hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}
    + \hat c_{q,\e}^\dag \hat c_{p,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}}
  = \f{U}{L} \sum_{p,q}
  \p{\hat c_{q,\g}^\dag \hat c_{q,\g} \hat c_{p,\e}^\dag \hat c_{p,\e}
    - \hat c_{q,\e}^\dag \hat c_{q,\g} \hat c_{p,\g}^\dag \hat c_{p,\e}},
\end{align}
where we can use the single-particle pseudo-spin-1/2 Pauli operators
in \eqref{eq:pseudospin} to write
\begin{align}
  H_{\t{int}} = \f{U}{L} \sum_{p,q}
  \sp{\p{\f{\1_q-\sigma_q^z}{2}} \p{\f{\1_p+\sigma_p^z}{2}}
    - \sigma_q^+ \sigma_p^-}.
\end{align}
Up to a global shift in energy, this result is equivalently
\begin{align}
  H_{\t{int}}
  = - \f14 \f{U}{L} \sum_{p,q} \v{\sigma}_q\c\v{\sigma}_p
  = - \f{U}{L} \sum_{p,q} \v s_q \c\v s_p
  = - \f{U}{L} \v S \c \v S,
\end{align}
where
\begin{align}
  \v{\sigma}_p
  \equiv \p{\sigma_p^x,\sigma_p^y,\sigma_p^z},
  &&
  \v s_p \equiv \p{s_p^x,s_P^y,s_p^z},
  &&
  \v S \equiv \f12 \sum_p \v{\sigma}_p.
  \label{eq:H_int}
\end{align}
For a fixed particle number $N$, the collective spin Hamiltonian
$H_{\t{int}}$ has electronic eigenstates $\set{\ket{Sm}}$ with total
(pseudo-)spin $S$ and projection $m$ onto a quantization axis; the
corresponding energies are $\bk{Sm|H_{\t{int}}|S,m}=-\p{U/L}S\p{S+1}$.
In particular, the ground-state $S=N/2$ manifold is spanned by the
Dicke states
\begin{align}
  \ket{m} \equiv \ket{N/2,m}
  = {N \choose N/2+m}^{-1/2} S_+^{N/2+m} \ket{\g}^{\otimes N},
  &&
  S_+ \equiv \sum_n \sigma_n^+,
  \label{eq:dicke_states}
\end{align}
where $n$ indexes an individual atom,
$\sigma_n^+\equiv\hat c_{n,\e}\hat c_{n,\g}$ is an individual
spin-raising operator, and $N/2+m=0,1,\cdots,N$ is the number of
electronic state excitations in the state $\ket{m}$
\cite{swallows2011suppression}.  In words, the Dicke state $\ket{m}$
is a uniform superposition of all states with $N/2+m$ total electronic
excitations.

@@

In total, free evolution of atoms on a lattice is governed by the
Hamiltonian
\begin{align}
  H_{\t{free}}
  = H_{\t{int}} + H_{\t{SOC}}
  = -\f{U}{L} \v S\c\v S - \sum_n h_n s_n^z,
\end{align}
for $n$ indexing an individual atom.  When the ``local'' fields $h_n$
are small in magnitude compared to the collective spin gap $\eta U$
and we initialize all atoms in the ground-state subspace of
$H_{\t{int}}$, we can treat the action of the spin-orbit coupling
Hamiltonian perturbatively, which yields the effective free-evolution
Hamiltonian (see Appendix \ref{sec:squeezing_derivation} or Peiru's
notes)
\begin{align}
  H_{\t{free,eff}} = -\bar h S_z + \chi S_z^2,
\end{align}
where
\begin{align}
  \bar h \equiv \f1N \sum_n h_n,
  &&
  \tilde h^2 \equiv \f1N \sum_n \p{h_n - \bar h}^2,
  &&
  \chi \equiv \f{\tilde h^2}{\p{N-1}\eta U}.
\end{align}
At temperatures $T\gg J_0$, the mean effective field $\bar h=0$, while
for $T\sim J_0$ the collective rotation due to the $S_z$ term can be
eliminated by use of a rotating frame.  The free-evolution Hamiltonian
thus simplifies further to the one-axis twisting Hamiltonian
\begin{align}
  H_{\t{OAT}}^z = \chi S_z^2.
  \label{eq:H_OAT}
\end{align}
An example set of parameters and values for one-axis twisting via
$H_{\t{OAT}}^z$ with strontium-87 in a 1-D magic-wavelength lattice is
provided in Table \ref{tab:parameters}, along with the optimal
one-axis twisting squeezing time
$t_{\t{opt}}^{\t{OAT}}\sim N^{-2/3}\chi^{-1}\sim\eta
N^{1/3}=N^{4/3}/L$.  Lattice depths are provided in units of the
lattice recoil energy $E_R$.

While one-axis twisting provides a spin fluctuation noise floor which
scales with atom number as $\sim1/N^{2/3}$, it is possible to beat
this limit and achieve a noise floor scaling as $\sim1/N$ via
so-called two-axis twisting.  We can transform the one-axis twisting
Hamitlonian in \eqref{eq:H_OAT} into an effective two-axis twisting
Hamiltonian by using a pulsed drive sequence developed in
ref.~\cite{liu2011spin}.  The necessary drive sequence requires the
pulsed application of a clock laser, which turns on a collective spin
Hamiltonian
\begin{align}
  H_{\t{clock}} = -\Omega \sum_n s_n^x = -\Omega S_x.
\end{align}
If $\abs{\Omega}\gg N\chi$, then we can apply $H_{\t{clock}}$ for
short times $\theta/\abs{\Omega}\ll\p{N\chi}^{-1}$ (during which we
can neglect $H_{\t{eff}}$) to realize collective rotations of the form
$\exp\p{\pm i\theta S_x}$.  Thus
\begin{enumerate*}[label=(\roman*)]
\item acting with $\exp\sp{i\p{\pi/2}S_x}$,
\item waiting for a time $2\tau$,
\item acting with $\exp\sp{-i\p{\pi/2}S_x}$, and
\item waiting for a time $\tau$
\end{enumerate*}
realizes the unitary
\begin{align}
  U_1 = \exp\p{-i\tau\chi S_z^2} \exp\p{-i\f{\pi}{2}S_x}
  \exp\p{-i2\tau\chi S_z^2} \exp\p{i\f{\pi}{2}S_x}
  = \exp\p{-i\tau\chi S_z^2} \exp\p{-i2\tau\chi S_y^2}.
\end{align}
If $\tau\chi\ll1$, then
\begin{align}
  U_1 \approx \exp\sp{-i\tau\chi\p{S_z^2 + 2S_y^2}}
  = \exp\sp{-i\tau\chi\p{S^2 + S_y^2 - S_x^2}}
  \simeq \exp\sp{-i\tau\chi\p{S_y^2 - S_x^2}},
\end{align}
where $\simeq$ denotes equality up to an overall phase, which holds
above because $S^2$ is conserved by all dynamics.  Repeating the above
sequence $n$ times for a time $t=3n\tau$ thus results in the unitary
\begin{align}
  U = \exp\sp{-in\tau\chi\p{S_y^2 - S_x^2}}
  = \exp\sp{-i\p{t/3}\chi\p{S_y^2 - S_x^2}},
\end{align}
which is equivalently generated by evolution for a time $t$ under the
two-axis countertwisting Hamiltonian
\begin{align}
  H_{\t{TAT}}^{y,x} = \f{\chi}{3}\p{S_y^2 - S_x^2}.
  \label{eq:H_TAT_pulse}
\end{align}
Despite the factor of 1/3 in the effective two-axis countertwisting
strength, the optimal squeezing time for this two-axis twisting
Hamiltonian is generally close to that for one-axis twisting with
$H_{\t{OAT}}^z$ in \eqref{eq:H_OAT} \cite{liu2011spin}.

\begin{figure}
  \centering \includegraphics{squeezing_comparison_TAT.pdf}
  \caption{Comparison of squeezing through $H_{\t{OAT}}^z$ and
    $H_{\t{TAT}}^{z,x}$ via the parameters in Table
    \ref{tab:parameters} and the continuous two-axis twisting protocol
    with $\omega=27\times2\pi$ Hz and $\beta=0.906$.}
  \label{fig:squeezing_comparison}
\end{figure}

\section{Effective spin squeezing in the presence of a weak drive}
\label{sec:squeezing_derivation}

Suppose we have a Hamiltonian of the form

@@

  H = H_0 + V,
  H_0 = - \f{U}{L} \v S\c\v S,
  V = - \sum_n h_n s_n^z - \Omega S_x,

@@

and we consider $N$-particle states initially in the ground-state
manifold $\G_0$ of $H_0$, which have total spin $S=N/2$.  If the
largest eigenvalue of $V$ is smaller in magnitude than half of the
collective spin gap $N\abs{U}/L=\eta\abs{U}$, i.e.~the energy gap
under $H_0$ between $\G_0$ and its orthogonal complement $\E_0$, then
we can formally develop a perturbative treatment for the action of $V$
on $\G_0$.  Such a treatment yields an effective Hamiltonian on $\G_0$
of the form $H_{\t{eff}}=\sum_pH_{\t{eff}}^{(p)}$, where
$H_{\t{eff}}^{(p)}$ is order $p$ in $V$.  Letting $\P_0$ ($\Q_0$) be a
projector onto $\G_0$ ($\E_0$) and $X$ denote any operator on
$\G_0\cup\E_0$ (i.e.~the entire Hilbert space), we define the
superoperators
  \D X \equiv \P_0 X \P_0 + \Q_0 X \Q_0,
  \O X \equiv \P_0 X \Q_0 + \Q_0 X \P_0,
which select the diagonal ($\D$) and off-diagonal ($\O$) parts of $X$
with respect to $\G_0$ and $\E_0$, and
  \L X \equiv \sum_{\alpha,\beta}
  \f{\op{\alpha}\O X\op{\beta}}{E_\alpha-E_\beta},
  \t{where}
  H_0 = \sum_\alpha E_\alpha \op\alpha.
The first few terms in the expansion of the effective Hamiltonian
$H_{\t{eff}}$ are then, as derived in
ref.~\cite{bravyi2011schrieffer},
  H_{\t{eff}}^{(0)} = \P_0 H_0 \P_0,
  H_{\t{eff}}^{(1)} = \P_0 V \P_0,
  H_{\t{eff}}^{(2)} = -\f12 \P_0 \sp{\O V,\L V} \P_0.
  \label{eq:H_eff_012}
If we add a constant to $H_0$ such that $E_\psi=0$ for all
$\ket\psi\in\G_0$, or equivalently we measure all energies relative to
that of the ground-state manifold $\G_0$ with respect to $H_0$, it
immediately follows that $H_{\t{eff}}^{(0)}=0$.  To calculate
$H_{\t{eff}}^{(1)}$, we note that the ground-state manifold $\G_0$ is
spanned by the Dicke states $\ket{m}$ defined in
\eqref{eq:dicke_states}, in terms of which we can expand the
collective spin-$z$ operator as $S_z=\sum_mm\op{m}$.  We can likewise
expand the collective spin-$x$ operator $S_x$ in terms of $x$-oriented
Dicke states $\ket{m_x}$ as $S_x=\sum_mm\op{m_x}$.  The ground-state
projector $\P_0$ onto $\G_0$ can be expanded in either basis as
$\P_0=\sum_m\op{m}=\sum_m\op{m_x}$.  Defining the mean and residual
fields

@@

  b_n \equiv h_n - \bar h,

@@

we can then write

@@

  V = -\sum_n \p{b_n+\bar h} s_n^z - \Omega S_x
  = -\sum_n b_n s_n^z - \bar h S_z - \Omega S_x,

@@

and in turn

@@

  H_{\t{eff}}^{(1)}
  = \P_0\p{-\sum_n b_n s_n^z - \bar h S_z - \Omega S_x} \P_0
  = -\sum_n b_n \P_0 s_n^z\P_0 - \bar h S_z - \Omega S_x,
where we used the fact that $\P_0 S_{j=z,x} \P_0 = S_j$.  By
construction, the residual fields are mean-zero, i.e.~$\sum_nb_n=0$.
Using the particle-exchange symmetry of the Dicke states, we can thus
expand
  \sum_n b_n \P_0 s_n^z \P_0
  = \sum_{n,m,m'} b_n \op{m} s_n^z \op{m'}
  = \sum_n b_n \sum_{m,m'} \op{m} s_1^z \op{m'}
  = 0,
  H_{\t{eff}}^{(1)} = - \bar h S_z - \Omega S_x.
To calculate the second-order effective Hamiltonian
$H_{\t{eff}}^{(2)}$, we let $\B_0\p{\E_0}$ denote an eigenbasis of
$H_0$ for the excited subspace $\E_0$, define the operator
  \I \equiv \sum_{\ket\alpha\in\B_0\p{\E_0}} \f{\op\alpha}{E_\alpha},
which sums over projections onto excited states with corresponding
energetic suppression factors, and expand
  \L X = \O\p{\L X}
  = \Q_0 \L X \P_0 + \P_0 \L X \Q_0
  = \I X \P_0 - \P_0 X \I.
The expression for $H_{\t{eff}}^{(2)}$ in \eqref{eq:H_eff_012} then
simplifies to
  H_{\t{eff}}
  = -\f12 \P_0 \p{\sp{\O V, \I V \P_0} - \sp{\O V, \P_0 V \I}} \P_0
  = -\P_0 V \I V \P_0.
The only part of $V$ which is off-diagonal with respect to the ground-
and excited-state manifolds $\G_0$ and $\E_0$ is $-\sum_nb_ns_n^z$,
and the individual operators in this term can only change the total
spin $S$ by at most 1.  It is therefore sufficient to expand $\I$ in a
basis for states with total spin $S=N/2-1$, which is provided by the
spin-wave states
  \ket{mk}
  \equiv \ket{N/2-1,m,k}
  \equiv {N-1 \choose \p{N/2-m}\p{N/2-m+1}}^{-1/2}
  \sum_{n=1}^N e^{2\pi ikn/N} s_n^+ \ket{N/2,m-1},
for $k=1,2,\cdots,N-1$ \cite{swallows2011suppression}.  Using the fact that all spin-$z$ operators preserve the projection of total spin onto the $z$ axis, we then have that
  H_{\t{eff}}^{(2)}
  = -\f1{\eta U} \sum_{m,m',k,n,n'} b_n b_{n'}
  \Bk{m| s_n^z \op{m'k} s_{n'}^z |m} \op{m}.
  \label{eq:general_H_eff_2}
The relevant matrix elements between the Dicke states and the
spin-wave states are \cite{swallows2011suppression}
  \bk{m|s_n^z|m'k}
  = e^{2\pi i k n/N} \sqrt{\f{(N/2)^2-m^2}{N^2 (N-1)}}~ \delta_{m,m'},

@@

  = -\f1{\eta U} \sum_m \f{(N/2)^2-m^2}{N^2 (N-1)} \op{m}
  \sum_{k,n,n'} b_n b_{n'} e^{2\pi ik\p{n-n'}/N}.

@@

  \sum_{k,n,n'} b_n b_{n'} e^{2\pi ik\p{n-n'}/N}
  = \sum_{n,n'} b_n b_{n'} \sum_{k=1}^{N-1} e^{2\pi ik\p{n-n'}/N}

@@

  \sum_{k,n,n'} b_n b_{n'} e^{2\pi ik\p{n-n'}/N}

@@

  H_{\t{free,eff}} = - \bar h S_z - \Omega S_x + \chi S_z^2,

@@
